Inferring stellar rotation periods using Gaussian processes

### 2016-11-30 documents/ms.tex
running on a kepler target and making python 3 compatible

\newcommand{\columbia}{1}
\newcommand{\oxford}{2}
\newcommand{\uw}{3}
\newcommand{\nlightcurves}{333}
\newcommand{\Kepler}{{\it Kepler}}
\newcommand{\kepler}{\Kepler}
\newcommand{\corot}{{\it CoRoT}}
\newcommand{\Ktwo}{{\it K2}}
\newcommand{\ktwo}{\Ktwo}
\newcommand{\TESS}{{\it TESS}}
\newcommand{\LSST}{{\it LSST}}
\newcommand{\SDSS}{{\it SDSS}}
\newcommand{\PLATO}{{\it PLATO}}
\newcommand{\Teff}{$T_{\mathrm{eff}}$}
\newcommand{\teff}{$T_{\mathrm{eff}}$}
\newcommand{\FeH}{[Fe/H]}
\newcommand{\feh}{[Fe/H]}
\newcommand{\ie}{{\it i.e.}}
\newcommand{\eg}{{\it e.g.}}
\newcommand{\logg}{log \emph{g}}
\newcommand{\dnu}{$\Delta \nu$}
\newcommand{\numax}{$\nu_{\mathrm{max}}$}

@@

\title{Inferring stellar rotation periods with Gaussian processes}

\author{%
   Ruth Angus\altaffilmark{\columbia}
   Suzanne Aigrain\altaffilmark{\oxford}
   \& Daniel Foreman-Mackey\altaffilmark{\uw}
}

\altaffiltext{\columbia}{Simons Fellow, Columbia University, NY, NY,
ruthangus@gmail.com}
\altaffiltext{\oxford}{Subdepartment of Astrophysics, University of Oxford,
OX1 3RH, UK}
\altaffiltext{\uw}{Department of Astronomy, University of Washington, Seattle,
WA}

@@

GP models parameterise the covariance between data points and a kernel
function provides the form of covariance matrix parameterisation.
For example, take the time-series in figure \ref{fig:GP_example}.
% This is a \kepler\ light curve of Earth-like planet host Kepler-452: a G-type
% star that rotates once every $\sim$ 15 days.
This is the \kepler\ light curve of KIC 1430163, a relatively active star that
rotates once every $\sim$ 4 days.
The stochastic variability in this time-series is typical of \kepler\ FGK
stars.
Clearly, data points in this light curve are correlated.
Points that are close together in time are tightly correlated and those that
are widely separated in time are loosely correlated.
The way in which the covariance varies with the separation between data points
is modelled when using GP regression; it is not the data but the {\it
covariance structure} that is modelled.
This fact is what gives GPs their flexibility---they can model any time
series with a similar covariance structure.
In addition, a very simple function can usually capture the covariance
structure of a light curve, whereas a much more complicated one may be
required to model the time-series itself.
% The light curve of \Kepler-452 has been modelled with a GP in figure
% \ref{fig:GP_example}, shown in orange.
In figure \ref{fig:GP_example} the light curve of KIC 1430163 has been
modelled with a GP.
% Both provide adequate fits to the data, however only the periodic kernel
% function, `QP' is a useful model because it has a periodic parameter.
% I return to this point shortly.

@@

In order to infer a stellar rotation period from a light curve, we fit a GP
model with a QP kernel function to the data.
The likelihood of the model, conditioned on the data could then be maximised
in order to find the maximum likelihood value for $P$.
In this study however, the full posterior PDFs of the parameters are explored
using MCMC.
This latter approach comes at a cost: a GP model is expensive to compute once
as it requires a matrix inversion and determinant evaluation, let alone the
many thousands of times that is necessary to fully explore the posteriors of
the parameters.
However, fully exploring the posterior PDF of $P$, allows for an accurate
estimate of the uncertainty on the rotation period.

\begin{figure}
\begin{center}
\includegraphics[width=6in, clip=true]{figures/001430163.pdf}
\caption[A light curve with a GP model.]
{Light curve of KIC 1430163, an active star with a rotation period of $\sim$ 4
days.
The blue solid line shows a fit to the data using
a Gaussian process model with a QP covariance kernel function and the pink
broken line shows a fit with a SE kernel function.
Both kernel functions provide an adequate fit to the data, however we use the
QP function because it has a useful parameter that corresponds to the star's
rotation period.}
\label{fig:GP_example}
\end{center}
\end{figure}

@@

In order to test the GP method we attempted to the measure rotation periods of
a set of simulated light curves.
We simulated \nlightcurves\ light curves using a spot model similar to that
used
% We used light curves simulated
for the \citet{Aigrain2015} `hare and hounds' rotation period recovery
experiment.
These light curves were simulated by placing dark, circular spots with slowly
evolving size, on the surface of bright, rotating spheres, ignoring
limb-darkening effects.
\citet{Aigrain2015} simulated one thousand light curves in order to test the
ability of participating teams to recover both the stellar rotation periods
and the rotational shear: the amplitude of surface differential rotation.
% Since we are not interested in recovering differential rotation we only used
% those light curves simulated {\it without} differential rotation, of which
% there are \nlightcurves.
Unlike in their study we are not interested in recovering differential
rotation in this work, so did not include it in our simulations.
We opted to use only solid-body rotators because differential rotation may
produce some additional scatter in the rotation period measurements recovered.
This code can also be adjusted to produce more realistic light curves by
altering spot lifetimes.
Stars with spot lifetimes that are long relative to their rotation periods
will have highly periodic light curves.
Those with short spot lifetimes will be more quasi-periodic.
We fixed the mean spot lifetime at an arbitrary value of 30.5 days for all
light curves in these initial tests.
In future we plan to include both differential rotation and variable spot
lifetimes in our light curve model.
Light curves were simulated with a real \Kepler\ long-cadence time array, with
one data point every thirty minutes over a four year duration.
The rotation periods of the simulations were randomly drawn from a log-uniform
distribution between 0.5-60 days.
Figure \ref{fig:noise-free_lc} shows an example of a simulated, noise-free
light curve with a period of 4.2 days.

% The ranges and distributions of the physical stellar parameters used in the
% simulated light curves are tabulated below, in table
% \ref{tab:simulation_parameters}.

% \begin{table*}
% \begin{center}
% \caption{Ranges and distributions of parameters used to simulate light curves
% in \citet{aigrain2015}}
% \begin{tabular}{lcc}
% \hline\hline
%     Parameter & Range & Distribution \\
%     \hline
%     Rotation period, $P_{rot}$ & 10 - 50 days (90\%) & log uniform \\
%     & 1 - 10 days (10\%) & log uniform \\
%     Activity cycle length & 1 - 10 years & log uniform \\
%     Inclination & 0 - 90$^\circ$ & Uniform in $\sin^2i$ \\
%     Decay timescale & (1 - 10) $\times P_{rot}$ & log uniform \\
% \hline
% \end{tabular}
% \end{center}
% \end{table*}
% \label{tab:simulation_parameters}

% and figure \ref{fig:period_hist} shows the
% distribution of rotation periods in the \citet{aigrain15} sample.
% \begin{figure}
% \begin{center}
% \includegraphics[width=6in, clip=true]{figures/acf_compare_noise-free_hist.pdf}
% \caption{A histogram of the rotation periods used to generate the simulated
% light curves in \citet{aigrain15}.}
% \label{fig:period_hist}
% \end{center}
% \end{figure}

\begin{figure}
\begin{center}
\includegraphics[width=6in, clip=true]{figures/noise-free_lc.pdf}
% \includegraphics[width=6in, clip=true]{figures/thesis_plot.pdf}
\caption[A simulated light curve.]
{An example simulated, noise-free light curve. This `star' has a
rotation period of 4.2 days.}
\label{fig:noise-free_lc}
\end{center}
\end{figure}

Initial tests were conducted on noise-free light curves, in order to provide
proof-of-concept.
We attempted to recover the rotation periods of these \nlightcurves\ light
curves
% , both with and without noise, (\ie before {\it and} after being injected
% into \kepler light curves)
using three rotation period recovery methods: the ACF method, the LS
periodogram method and the GP method.

@@

We calculated an ACF for each light curve using the method of
\citet{Mcquillan2013}.
In this method, an ACF is calculated for each light curve and smoothed by
convolving with a Gaussian with $\sigma=9$ days.
A rotation period is estimated as the lag-time of the first peak in the ACF,
unless the second peak is larger in which case {\it that} lag-time is
interpreted as the true period.
The second peak in an ACF can be larger than the first if there are two active
regions at or near opposite longitudes on the surface of the star---this would
produce a light curve with two dips per rotation period.
% By extension, if three active longitudes existed at 60$^\circ$ separations on
% the stellar surface, the first peak in the ACF may be at one third of the true
% rotation period and so on.
An example ACF of the light curve in figure~\ref{fig:noise-free_lc} is shown
in figure \ref{fig:ACF_example}.

\begin{figure*}
\begin{center}
\includegraphics[width=6in, clip=true]{figures/noise-free_acf.pdf}
\caption[An ACF of a simulated light curve.]
{The autocorrelation function of the simulated, noise-free light
curve depicted in figure \ref{fig:noise-free_lc}.}
\label{fig:ACF_example}
\end{center}
\end{figure*}

The ACF method has proven to be extremely useful for measuring rotation
periods.
The catalogue of rotation periods of \Kepler\ stars provided in
\citet{Mcquillan2013} has been widely used by the community and has provided
ground-breaking results for stellar and exoplanetary science.
The results of the ACF method as tested in \citet{Aigrain2015} were positive
(see, for example their figure 8) as it produced a large number of accurate
rotation period measurements.
Another advantage of the ACF method is that it is conveniently fast to
implement.

We applied the ACF method to our sample of \nlightcurves\ noise-free,
simulated light curves.
Periods measured using the ACF method are plotted against the original
rotation period values used to generate the light curves in figure
\ref{fig:compare_noise_free}.
% 73\% of the injected rotation periods were recovered with a value lying
% within 10\% of the truth and 83\% within 20\%.
A noteworthy feature of this figure is that many of the points fall below the
$x=y$ line, \ie\ the recovered rotation periods are a little shorter than the
true periods.
We believe this is a feature of the peak position measurement method that is
performed on the ACF.
ACFs of stellar light curves can be roughly described as a cosine function
super-imposed on top of a decaying exponential.
In such a function the peak positions can be shifted towards the left--the
short period end--because the decaying exponential raises the left side of
each peak more than the right.
It is possible to model this effect of course, however the standard practice
is to simply measure the peak position without taking it into account.
We are still investigating the origin and implications of this effect.
% To demonstrate that this effect reproduces the underestimated periods seen in
% figure \ref{fig:compare_noise_free}, we simulated 1000 ACFs by generating
% cosine plus exponential functions with a range of periods, decay timescales
% and relative amplitudes.
% The measured position of the first peak in each simulated ACF is compared to
% the true position in figure \ref{fig:exp_sine_test}.
% The same underestimation of the peak position is apparent in this figure.
% Dashed lines show harmonics of $2P_{\mathrm{rot}}$ and $1/2P_{\mathrm{rot}}$.
% In several cases, twice the true rotation period is measured.
% The RMS of residuals is 1.59 days.

@@

\begin{figure*}
\begin{center}
% \includegraphics[width=6in, clip=true]{figures/acf_compare_noise-free.pdf}
\includegraphics[width=6in, clip=true]{figures/compare_acf.pdf}
\caption[ACF results.]
{Rotation periods measured using the ACF method as a function of the
`true' period used to generate \nlightcurves\ simulated, noise-free light
curves.}
\label{fig:compare_noise_free}
\end{center}
\end{figure*}

@@

For each simulated light curve, a Lomb-Scargle (LS) periodogram
\citep{Lomb1976, Scargle1982} was computed over a grid of 10,000 periods,
evenly spaced in frequency, between 1 and 100 days.
The period of the highest peak in the periodogram was adopted as the rotation
period.
The resulting recovered rotation periods are plotted as a function of the true
periods in figure \ref{fig:pgram_compare_noise_free}.
These recovered rotation periods are in general more accurate than the ACF
results: they do not systematically over or under predict rotation period.
They are however less precise than the ACF results---the 8.03 day RMS of
residuals is five times larger.
The longer-period measurements provide the dominant contribution to this RMS.
% We believe that the unsuccessful recoveries occurring preferentially for
% shorter rotation periods are produced by additional, longer timescale
% variations in some of the light curves.
% Specifically, by fluctuations produced by changes in the total spot coverage
% on the stellar surface.
% These fluctuations can occasionally be larger in magnitude than those produced
% by the rotational signal itself and are present in the majority of outlying
% light curves that occupy the top left of figure
% \ref{fig:pgram_compare_noise-free}.

\begin{figure*}
\begin{center}
\includegraphics[width=6in, clip=true]{figures/compare_pgram.pdf}
\caption[LS periodogram results.]
{Rotation periods measured using the LS periodogram method as a
function of the `true' rotation period used to generate \nlightcurves\
simulated light curves.}
\label{fig:pgram_compare_noise_free}
\end{center}
\end{figure*}

@@

We use the ACF period as an initial guess for the rotation period (this
decision is discussed further in section \textsection \ref{sec:discussion}).
We use a uniform prior over rotation period with bounds described below and
assert that the covariance decay timescale parameter, $l$ must be greater than
the rotation period.
This represents our assumption that the evolution timescale of active regions
is greater than stellar rotation periods.
This assumption may be more valid for late spectral types---hot stars have
smaller active regions which are likely to be shorter-lived.
For the remaining hyperparameters we use the following initial values and
log-uniform prior distributions:
\begin{eqnarray}
 	&	A_{initial} = e^{-5}, \sim \exp(U[-20:20]) \\ \nonumber
 	&	l_{initial} = e^{7}, \sim \exp(U[-20:20]), l<P \\ \nonumber
 	&	\Gamma_{initial} = e^{0.6}, \sim \exp(U[-20:20]) \\ \nonumber
 	&	\sigma_{initial} = e^{-16}, \sim \exp(U[-20:20]) \\ \nonumber
 	&	P_{initial} = P_{ACF}, \sim U([1 - 0.4]P_{ACF}<P_{ACF}<[1 +
0.4]P_{ACF}).
 \end{eqnarray}
 \label{eq:initialisation}
$\sigma$ is an additional white noise term added to the diagonal elements of
the covariance matrix. It is the fraction by which the observational
uncertainties have been underestimated.
If the errorbars reported on the data are too small, this parameter will be
non-zero.
In practice this parameter should always be non-zero when performing GP
regression.
This is because the covariance matrix must be positive definite, however
matrix inversion performed using most solvers is approximate, not exact,
therefore slight deviations from positive definitism can arise.
Including a small amount of extra variance in the model allows enough
flexibility that the matrix inversion algorithms do not run into numerical
issues.

We subsampled the light curves in order to reduce computation time.
The subsampling is altered for each light curve and depends, again, on the ACF
period estimate.
We found that retaining only 20 data points per rotation period, based on the
ACF period estimate, significantly reduced computation time whilst maintaining
performance.
The {\tt george} \citep{George} python package was used to implement the GP
model which uses the fast matrix solver, HODLR \citep{Ambikasaran2014}.
Matrix operations performed with HODLR scale as $N \log ^2 (N)$, where $N$ is
the number of data points.
We used {\tt emcee} \citep{Foreman-Mackey2013}, an affine invariant ensemble
sampler to explore the posterior PDFs of the model parameters.
The rotation period was taken as the median value of the posterior PDF.
The resulting measured rotation periods are compared to the true rotation
periods in figure \ref{fig:GP_compare_noise_free}.
The root-mean-square (RMS) of residuals for the GP results is 0.26 days.
The results of all three methods are plotted on the same axes in figure
\ref{fig:compare_noise_free}.

% The subsampling reduces computation time simply by contracting the data

\begin{figure*}
\begin{center}
\includegraphics[width=6in, clip=true]{figures/compare_gp.pdf}
\caption[GP results.]
{Measured vs true rotation period for \nlightcurves\ simulated light
curves using the GP method.}
\label{fig:GP_compare_noise_free}
\end{center}
\end{figure*}

\begin{figure*}
\begin{center}
\includegraphics[width=6in, clip=true]{figures/compare2.pdf}
\caption[All results.]
{Measured vs true rotation period for \nlightcurves\ simulated light
curves from the three different methods described in the text.}
\label{fig:compare_noise_free}
\end{center}
\end{figure*}

@@

% \begin{figure*}
% \begin{center}
% \includegraphics[width=6in, clip=true]{figures/prediction.pdf}
% \caption{A noisy simulated light curve with a rotation period of 17.04 days.
% The blue curve shows the GP model with the best-fit parameters.}
% \end{center}
% \end{figure*}
% \label{fig:compare_noise_free}

>>>>

\newcommand{\columbia}{1}
\newcommand{\oxford}{2}
\newcommand{\uw}{3}
\newcommand{\princeton}{4}
\newcommand{\nlightcurves}{333}
\newcommand{\Kepler}{{\it Kepler}}
\newcommand{\kepler}{\Kepler}
\newcommand{\corot}{{\it CoRoT}}
\newcommand{\Ktwo}{{\it K2}}
\newcommand{\ktwo}{\Ktwo}
\newcommand{\TESS}{{\it TESS}}
\newcommand{\LSST}{{\it LSST}}
\newcommand{\SDSS}{{\it SDSS}}
\newcommand{\PLATO}{{\it PLATO}}
\newcommand{\Teff}{$T_{\mathrm{eff}}$}
\newcommand{\teff}{$T_{\mathrm{eff}}$}
\newcommand{\FeH}{[Fe/H]}
\newcommand{\feh}{[Fe/H]}
\newcommand{\ie}{{\it i.e.}}
\newcommand{\eg}{{\it e.g.}}
\newcommand{\logg}{log \emph{g}}
\newcommand{\dnu}{$\Delta \nu$}
\newcommand{\numax}{$\nu_{\mathrm{max}}$}

@@

\title{Inferring stellar rotation periods using Gaussian processes}

\author{%
   Ruth Angus\altaffilmark{\columbia},
   Suzanne Aigrain\altaffilmark{\oxford},
   Timothy Morton\altaffilmark{\princeton}
   \& Daniel Foreman-Mackey\altaffilmark{\uw}
}

\altaffiltext{\columbia}{Simons Fellow, Columbia University, NY, NY,
ruthangus@gmail.com}
\altaffiltext{\oxford}{Subdepartment of Astrophysics, University of Oxford,
OX1 3RH, UK}
\altaffiltext{\uw}{Department of Astronomy, University of Washington, Seattle,
WA}
\altaffiltext{\princeton}{Department of Astronomy, Princeton University}

@@

GP models parameterise the covariance between data points and a kernel
function provides the form of covariance matrix parameterisation.
For example, take the time-series in figure \ref{fig:GP_example}.
This is the \kepler\ light curve of KIC 1430163, a relatively active star that
rotates once every $\sim$ 4 days.
The stochastic variability in this time-series is typical of \kepler\ FGK
stars.
Clearly, data points in this light curve are correlated.
Points that are close together in time are tightly correlated and those that
are widely separated in time are loosely correlated.
The way in which the covariance varies with the separation between data points
is modelled when using GP regression; it is not the data but the {\it
covariance structure} that is modelled.
This fact is what gives GPs their flexibility---they can model any time
series with a similar covariance structure.
In addition, a very simple function can usually capture the covariance
structure of a light curve, whereas a much more complicated one may be
required to model the time-series itself.
% The light curve of \Kepler-452 has been modelled with a GP in figure
% \ref{fig:GP_example}, shown in orange.
In figure \ref{fig:GP_example} the light curve of KIC 1430163 has been
modelled with a GP.
% Both provide adequate fits to the data, however only the periodic kernel
% function, `QP' is a useful model because it has a periodic parameter.
% I return to this point shortly.

@@

In order to infer a stellar rotation period from a light curve, we fit a GP
model with a QP kernel function to the data.
The likelihood of the model, conditioned on the data could then be maximised
in order to find the maximum likelihood value for $P$.
In this study however, the full posterior PDFs of the parameters are explored
using MCMC.
This approach comes at a cost: a GP model is expensive to compute once as it
requires a matrix inversion and determinant evaluation, let alone the many
thousands of times that is necessary to fully explore the posteriors of the
parameters.
However, fully exploring the posterior PDF of $P$ allows for an accurate
estimate of the uncertainty on the rotation period.

\begin{figure}
\begin{center}
\includegraphics[width=6in, clip=true]{figures/001430163.pdf}
\caption[A light curve with a GP model.]
{Light curve of KIC 1430163, an active star with a rotation period of $\sim$ 4
days.
The blue line shows a fit to the data using
a Gaussian process model with a QP covariance kernel function.}
\label{fig:GP_example}
\end{center}
\end{figure}

@@

In order to test the GP method we attempted to the measure rotation periods of
a set of simulated light curves.
We used \nlightcurves\ light curves generated for the \citet{Aigrain2015}
`hare and hounds' rotation period recovery experiment.
These light curves were simulated by placing dark, circular spots with slowly
evolving size, on the surface of bright, rotating spheres, ignoring
limb-darkening effects.
\citet{Aigrain2015} simulated one thousand light curves in order to test the
ability of participating teams to recover both the stellar rotation periods
and the rotational shear: the amplitude of surface differential rotation.
% Since we are not interested in recovering differential rotation we only used
% those light curves simulated {\it without} differential rotation, of which
% there are \nlightcurves.
For the purpose of demonstrating that our method can recover rotation periods,
we selected only the \nlightcurves\ out of the 1000 without differential
rotation.
We opted to use only solid-body rotators because differential rotation may
produce some additional scatter in the rotation period measurements recovered.
In future we intend to test whether we can recover differential rotation using
the GP method and will then use the full set of 1000 light curves.
The distributions of input parameters used to generate these light curves are
listed in \citet{Aigrain2015}.
Light curves were simulated with a real \Kepler\ long-cadence time array:
one data point every thirty minutes over a four year duration.
90\% of the rotation periods of the simulations were randomly drawn from a
log-uniform distribution between 10 and 50 days and 10\% from a log-uniform
distribution between 1 and 10 days.
The light curves were also generated with a range of stellar inclination
angles, activity levels, spot lifetimes and more.
Once simulated, the light curves were added to real \kepler\ light curves with
no obvious astrophysical variability in order to preserve the noise properties
of the time series.
Figure \ref{fig:demo_lc} shows an example of a simulated, noise-free
light curve with a period of 17.04 days.

The ranges and distributions of the physical stellar parameters used in the
simulated light curves are tabulated below, in table
\ref{tab:simulation_parameters}.

\begin{table*}
\begin{center}
\caption{Ranges and distributions of parameters used to simulate light curves
in \citet{aigrain2015}}
\begin{tabular}{lcc}
\hline\hline
    Parameter & Range & Distribution \\
    \hline
    Rotation period, $P_{rot}$ & 10 - 50 days (90\%) & log uniform \\
    & 1 - 10 days (10\%) & log uniform \\
    Activity cycle length & 1 - 10 years & log uniform \\
    Inclination & 0 - 90$^\circ$ & Uniform in $\sin^2i$ \\
    Decay timescale & (1 - 10) $\times P_{rot}$ & log uniform \\
\hline
\end{tabular}
\end{center}
\end{table*}
\label{tab:simulation_parameters}

and figure \ref{fig:period_hist} shows the
distribution of rotation periods in the \citet{aigrain15} sample.
\begin{figure}
\begin{center}
\includegraphics[width=6in, clip=true]{figures/period_hist.pdf}
\caption{A histogram of the rotation periods used to generate the simulated
light curves in \citet{aigrain15}.}
\label{fig:period_hist}
\end{center}
\end{figure}

\begin{figure}
\begin{center}
\includegraphics[width=6in, clip=true]{figures/demo_lc.pdf}
\caption[A simulated light curve.]
{An example simulated light curve. This `star' has a rotation period of 20.8
days.}
\label{fig:demo_lc}
\end{center}
\end{figure}

We attempted to recover the rotation periods of these \nlightcurves\ light
curves using three rotation period recovery methods: the ACF method, the LS
periodogram method and the GP method.

@@

We calculated an ACF for each light curve using the method of
\citet{Mcquillan2013}.
In this method, an ACF is calculated for each light curve and smoothed by
convolving with a Gaussian with $\sigma=9$ days.
A rotation period is estimated as the lag-time of the first peak in the ACF,
unless the second peak is larger in which case {\it that} lag-time is
interpreted as the true period.
The second peak in an ACF can be larger than the first if there are two active
regions at or near opposite longitudes on the surface of the star---this would
produce a light curve with two dips per rotation period.
An example ACF of the light curve in figure~\ref{fig:demo_lc} is shown
in figure \ref{fig:ACF_example}.

\begin{figure}
\begin{center}
\includegraphics[width=6in, clip=true]{figures/demo_ACF.pdf}
\caption[ACF of a simulated light curve.]
{An autocorrelation function of the simulated light curve shown in figure
\ref{fig:demo_lc}.
The vertical blue line shows the period measured using the ACF method (19.5
days) and the pink dashed line shows the period that was used to simulate the
light curve (20.8 days).}
\label{fig:demo_lc}
\end{center}
\end{figure}

The ACF method has proven to be extremely useful for measuring rotation
periods.
The catalogue of rotation periods of \Kepler\ stars provided in
\citet{Mcquillan2013} has been widely used by the community and has provided
ground-breaking results for stellar and exoplanetary science.
The results of the ACF method as tested in \citet{Aigrain2015} were positive
(see, for example their figure 8) as it produced a large number of accurate
rotation period measurements.
Another advantage of the ACF method is that it is fast to implement.

We applied the ACF method to the sample of \nlightcurves\ simulated light
curves.
Periods measured using the ACF method are plotted against the original
rotation period values used to generate the light curves in figure
\ref{fig:compare_noise_free}.
% 73\% of the injected rotation periods were recovered with a value lying
% within 10\% of the truth and 83\% within 20\%.
A noteworthy feature of this figure is that many of the points fall below the
$x=y$ line, \ie\ the recovered rotation periods are a little shorter than the
true periods.
This is a feature of the peak position measurement method that is performed on
the ACF.
ACFs of stellar light curves are periodic functions added to decaying
exponentials.
In such functions the peak positions can be shifted towards the left--the
short period end--because the decaying exponential raises the left side of
each peak more than the right.
It is possible to model this effect, however the standard practice is to
simply measure the peak position without taking it into account.
Additionally, the presence of gaps in the data may be contributing to this
effect.
Although, in our ACF method, gaps are interpolated over, these gaps still
affect the ACF by reducing its peak heights.
This amplifies the effect described above, in which the period is
underestimated due to the exponential decay in the ACF.
% To demonstrate that this effect reproduces the underestimated periods seen in
% figure \ref{fig:compare_noise_free}, we simulated 1000 ACFs by generating
% cosine plus exponential functions with a range of periods, decay timescales
% and relative amplitudes.
% The measured position of the first peak in each simulated ACF is compared to
% the true position in figure \ref{fig:exp_sine_test}.
% The same underestimation of the peak position is apparent in this figure.
% Dashed lines show harmonics of $2P_{\mathrm{rot}}$ and $1/2P_{\mathrm{rot}}$.
% In several cases, twice the true rotation period is measured.
% The RMS of residuals is 1.59 days.

@@

\begin{figure*}
\begin{center}
\includegraphics[width=6in, clip=true]{figures/compare_acf.png}
\caption[ACF results.]
{The `true' rotation periods used to generate \nlightcurves\ simulated light
curves vs the rotation periods measured using the ACF technique.}
\label{fig:compare_acf}
\end{center}
\end{figure*}

@@

For each simulated light curve, a Lomb-Scargle (LS) periodogram
\citep{Lomb1976, Scargle1982} was computed over a grid of 10,000 periods,
evenly spaced in frequency, between 1 and 100 days.
The period of the highest peak in the periodogram was adopted as the rotation
period.
The resulting recovered rotation periods are plotted as a function of the true
periods in figure \ref{fig:pgram_compare}.
These recovered rotation periods are in general more accurate than the ACF
results: they do not systematically over or under predict rotation period.
They are however less precise than the ACF results---the 8.03 day RMS of
residuals is five times larger.
The longer-period measurements provide the dominant contribution to this RMS.
% We believe that the unsuccessful recoveries occurring preferentially for
% shorter rotation periods are produced by additional, longer timescale
% variations in some of the light curves.
% Specifically, by fluctuations produced by changes in the total spot coverage
% on the stellar surface.
% These fluctuations can occasionally be larger in magnitude than those produced
% by the rotational signal itself and are present in the majority of outlying
% light curves that occupy the top left of figure
% \ref{fig:pgram_compare_noise-free}.

\begin{figure*}
\begin{center}
\includegraphics[width=6in, clip=true]{figures/compare_pgram.png}
\caption[LS periodogram results.]
{The `true' rotation periods used to generate \nlightcurves\ simulated light
curves vs the rotation periods measured using a LS periodogram technique.}
\label{fig:pgram_compare}
\end{center}
\end{figure*}

@@

We use the ACF period as an initial guess for the rotation period (this
decision is discussed further in section \textsection \ref{sec:discussion}).
We use Gausssian distributions in the natural logarithm of all the parameters
of our GP model.
The means and variances of the priors used were learned during the development
of this method.
In early stages of development we used very broad priors on all the parameters
uniform in log-space between -20 and 20.
We then examined the distributions of the model parameters for `sucessful' and
`unsuccessful' rotation period recoveries, where a recovery was termed
successful if it lay within 10\% of the injected period.
For the successful cases, the parameter distributions were narrow Gaussians.
The parameter distributions of the unsuccessful cases were broad Gaussians
with similar means.
We adopted the means and standard deviations of the narrow Gaussians
describing the parameter distributions of the successful cases to describe our
Gaussian priors.
The means and standard deviations of these $\ln$ Gaussian priors are given in
table \ref{tab:priors}.

\begin{table*}
\begin{center}
\caption{Means and standard deviations of the Gaussian priors on the natural
logarithms of the GP model parameters.}
\begin{tabular}{lcc}
Parameter & Mean & Standard Deviation \\
    \hline
    $A$ & -13 & 2.7 \\
    $l$ & 6.2 & 1.5 \\
    $\Gamma$ & -1.4 & 1.5 \\
    Rotation period, $P_{rot}$ & ACF period & ACF period $\times 2$ \\
    $\sigma$ & -17 & 5 \\
\end{tabular}
\end{center}
\end{table*}
\label{tab:priors}
The parameter positions were also initialised at the means of these
distributions during MCMC posterior exploration.

In addition, we set the following bounds on the rotation period and the $l$
parameter.
\begin{eqnarray}
 	&	\ln(0.5) < \ln(P_{\mathrm{rot}}) < \ln(T/2) \\ \nonumber
    &   0 < \ln(l) \\
\end{eqnarray}
\label{eq:priors}
Where $T$ is the total timespan of the light curve.
We limit rotation periods to be greater than half a day because we are chiefly
interested in rotation of FGK stars which are likely to be greater than this
lower limit, however this restriction could be relaxed in principle.
We also limit the rotation periods to be less than half of the total timespan
of the data since this is the longest period we expect to reliable recover.
We prevent $\ln(l)$ from falling beneath zero both because we expect that
astrophysical signals in light curves should have longer correlation
timescales than this and in practice we found that if $l$ is unrestricted, it
can occasionally wander towards a small value, resulting in an incorrect
rotation period estimate.

$\sigma$ is an additional white noise term added to the diagonal elements of
the covariance matrix. It is the fraction by which the observational
uncertainties have been underestimated.
If the uncertainties reported on the data are too small, this parameter will
be non-zero.
In practice however, this parameter should always be non-zero when performing
GP regression.
This is because the covariance matrix must be positive definite, however
matrix inversion performed using most solvers is approximate, not exact,
therefore slight deviations from positive definiteness can arise.
Including a small amount of extra variance in the model allows enough
flexibility that the matrix inversion algorithms do not run into numerical
issues.

We subsampled the light curves in order to reduce computation time.
We randomly subsampled the data, retaining an average of four data points per
day.
This allowed us to recover periods down to half a day --- the minimum
timescale of interest for this project, whilst reducing computational costs.
The {\tt george} \citep{George} python package was used to implement the GP
model which uses the fast matrix solver, HODLR \citep{Ambikasaran2014}.
Matrix operations performed with HODLR scale as $N \log ^2 (N)$, where $N$ is
the number of data points.
We used {\tt emcee} \citep{Foreman-Mackey2013}, an affine invariant ensemble
sampler to explore the posterior PDFs of the model parameters.
The rotation period was taken as the sample with the maximum likelihood value.
The resulting measured rotation periods are compared to the true rotation
periods in figure \ref{fig:compare_mcmc}.

@@
